---
title: "Chris's Predictions"
output: html_document
date: "2024-10-15"
---

## Introduction

Below I clean and prepare the data. I ended up just taking the weekly average. I limit the analysis to Trump and Harris, and all polls after 7/21. 

```{r}
library(dplyr)
library(lubridate)
library(knitr)


knitr::opts_chunk$set(echo = TRUE)
poll <- read.csv("~/Dropbox/github_repos/pol683/electionPrediction/presidentialPredictions24/president_polls.csv") %>%
  filter(mdy(end_date) > as.Date("07/21/24", "%m/%d/%y")) 

# Convert all dates to workweeks
dates = poll$end_date %>% mdy()  
workweek_start = floor_date(dates, unit = "week", week_start = 1)
day_start = floor_date(dates, unit = "day")
poll$day_id = as.integer(factor(day_start))  - 1
poll$week = as.integer(factor(workweek_start))  - 1
poll$week_id = workweek_start

poll = poll %>% group_by(state) %>% arrange(end_date) %>%
  filter(answer %in% c("Harris", "Trump"))  %>%
  group_by(day_id, state, candidate_name, end_date) %>%
  filter(state != "") %>%
  summarize(
    poll_average = mean(pct, na.rm = TRUE)
  ) %>%
  arrange(state, candidate_name)

poll %>% head()

write.csv(poll, file = "~/Dropbox/github_repos/pol683/electionPrediction/presidentialPredictions24/shiny/predictions/predictions/poll.csv")
```

## States over time


The statistical model I used to estimate our predictions is this:

```{r}
poll %>% 
ggplot(aes(x = as.Date(end_date,  "%m/%d/%y" ), y = poll_average, color = candidate_name)) +
  geom_jitter(size = 3, alpha = 0.3 ) +
  geom_smooth(method = "loess", se = FALSE) +  # Add trend lines
  labs(title = "Poll Averages",
       x = "Day",
       y = "Poll Average") +
  theme_minimal() +
  theme(legend.position = "none") + 
    scale_color_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red")) 
  # add dates using end_da
  
```

## BRMS

Estimate two models, one for Harris, one for Trump. This type of model is called a latent growth model. Parameters vary across states.

```{r, echo  = FALSE, message = FALSE, warning = FALSE}
library(brms)
    model2 <- brms::brm(
      poll_average ~ day_id + I(day_id^2) + I(day_id^3) +  (1+day_id + I(day_id^2) + I(day_id^3)|state),
      data = poll %>% filter(candidate_name == "Donald Trump"),
      family = gaussian(),
      iter = 1000,
      chains = 4,
      cores = getOption("mc.cores", 10),
    )
    model1 <- brms::brm(
      poll_average ~ day_id + I(day_id^2) + I(day_id^3) +  (1+day_id + I(day_id^2) + I(day_id^3)|state),
      data = poll %>% filter(candidate_name == "Kamala Harris"),
      family = gaussian(),
      iter = 1000,
      chains = 4,
      cores = getOption("mc.cores", 10),
    )
model1
model2

list(harris = model1, trump =  model2) -> models

save(models, file = "~/Dropbox/github_repos/pol683/electionPrediction/presidentialPredictions24/shiny/predictions/predictions/models.rda")


```

Here are predictions for both candidates

```{r}
bind_rows(
poll %>%
      tidybayes::add_epred_draws(model2) %>%
      group_by(day_id, state) %>%
      summarize(
        .value = mean(.epred),
        .lower = quantile(.epred, 0.025),
        .upper = quantile(.epred, 0.975)
      ) %>%
  mutate(candidate_name = "Kamala Harris"),
poll %>%
      tidybayes::add_epred_draws(model1) %>%
      group_by(day_id, state) %>%
      summarize(
        .value = mean(.epred),
        .lower = quantile(.epred, 0.025),
        .upper = quantile(.epred, 0.975)
      ) %>%
  mutate(candidate_name = "Donald Trump")
) %>%
ggplot(aes(x = day_id, y = .value, group = state, fill = state)) +
  facet_wrap(~candidate_name, scales = "free_y") + 
  geom_line(aes(color = state), size = 1, alpha = 0.3 ) +
  geom_ribbon(aes(ymin = .lower, ymax = .upper), alpha = 0.1) +
  labs(title = "Trajectories",
       x = "Week",
       y = "Poll Average") +
  theme_minimal() +
  theme(legend.position = "none") +
  # range y
  scale_y_continuous(limits = c(25, 75))
```


## Predictions


```{r}
my_predictions = 
                    bind_rows(
                    poll %>%
                      group_by(state) %>%
                      #extract max day and add 1
                      summarize(day_id = max(day_id) + 1) %>%
                      tidybayes::add_epred_draws(model2) %>%
                      group_by(state) %>%
                      # summarize posterior
                      summarize(
                        .value = mean(.epred),
                        .lower = quantile(.epred, 0.025),
                        .upper = quantile(.epred, 0.975)
                      ) %>%
                        mutate(candidate_name = "Donald Trump"),
                    poll %>%
                      group_by(state) %>%
                      #extract max day and add 1
                      summarize(day_id = max(day_id) + 1) %>%
                      tidybayes::add_epred_draws(model1) %>%
                      group_by(state) %>%
                      # summarize posterior
                      summarize(
                        .value = mean(.epred),
                        .lower = quantile(.epred, 0.025),
                        .upper = quantile(.epred, 0.975)
                      ) %>%
                      mutate(candidate_name = "Kamala Harris") 
) %>%
  filter(!stringr::str_detect(state, "CD"))
```

Here are my predictions:
```{r}
# Drop states with large margins; polling is sporadic, and increases the confidence intervals
my_predictions = my_predictions %>%
  select(state, candidate_name, .value) %>%
  tidyr::spread(candidate_name, .value) %>%
  mutate(margin = `Donald Trump` - `Kamala Harris`) %>%
  dplyr::right_join(my_predictions, by = "state") %>%
  filter(abs(margin) < 10)
head(my_predictions)

```

Here is the total United States, territories, and the District of Columbia, to see what's missing.
```{r}
### move to global
electoral_votes = read.csv("~/Dropbox/github_repos/pol683/electionPrediction/presidentialPredictions24/shiny/predictions/predictions/electoral_votes.csv")
all_states_and_territories <- c(
  "Alabama", "Alaska", "American Samoa", "Arizona", "Arkansas", "California", 
  "Colorado", "Connecticut", "Delaware", "District of Columbia", "Florida", 
  "Georgia", "Guam", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", 
  "Kentucky", "Louisiana", "Maine", "Maryland", "Massachusetts", "Michigan", 
  "Minnesota", "Mississippi", "Missouri", "Montana", "Nebraska", "Nevada", 
  "New Hampshire", "New Jersey", "New Mexico", "New York", "North Carolina", 
  "North Dakota", "Northern Mariana Islands", "Ohio", "Oklahoma", "Oregon", 
  "Pennsylvania", "Puerto Rico", "Rhode Island", "South Carolina", "South Dakota", 
  "Tennessee", "Texas", "Utah", "Vermont", "Virgin Islands", "Virginia", 
  "Washington", "West Virginia", "Wisconsin", "Wyoming"
)

# Create a second vector including the remaining U.S. states and territories
remaining_states_and_territories <- dplyr::setdiff(all_states_and_territories, 
                                                   with(my_predictions, unique(state)
))

trump_states <- c("Alabama", "Alaska", "Arkansas", "Idaho", "Indiana", "Kansas", 
                  "Kentucky", "Louisiana", "Mississippi", "Missouri", "Montana", 
                  "Nebraska", "North Dakota", "Oklahoma", "South Carolina", 
                  "South Dakota", "Tennessee", "Utah", "West Virginia", 
                  "Wyoming")

harris_states <- c("California", "Colorado", "Connecticut", "Delaware", 
                   "District of Columbia", "Hawaii", "Illinois", "Maine", 
                   "Maryland", "Massachusetts", 
                   "New Hampshire", "New Jersey", "New Mexico", 
                   "New York", "Oregon", "Rhode Island", "Vermont", 
                   "Washington")

# Create a data frame with states and their voting preferences
state_preferences <- data.frame(
  state = all_states_and_territories,
  prediction = ifelse(all_states_and_territories %in% trump_states, "Donald Trump", 
                          ifelse(all_states_and_territories %in% harris_states, "Kamala Harris", NA)),
  stringsAsFactors = FALSE
)

# Print the data frame
states = state_preferences  %>%
  left_join(my_predictions, by = "state")  %>%
   mutate(dif = plogis(`Donald Trump`/100 - `Kamala Harris`/100))  %>%
   select(state, candidate_name, .value, dif, prediction)   %>%
   pivot_wider(names_from = candidate_name, values_from = `.value`) %>%
   inner_join(electoral_votes, by = "state") 

for(i in 1:nrow(states)){
  if(is.na(states$prediction[i])){
    states$prediction[i] = ifelse(states$`Donald Trump`[i] > states$`Kamala Harris`[i], "Donald Trump", "Kamala Harris")
  }}

states %>% 
  mutate(
    electoral_Trump = ifelse(prediction == "Donald Trump", electoral_votes, 0),
    electoral_Harris = ifelse(prediction == "Kamala Harris", electoral_votes, 0)
  ) -> states

 write.csv(states, file = "~/Dropbox/github_repos/pol683/electionPrediction/presidentialPredictions24/shiny/predictions/predictions/electoral_predictions.csv")


```









